---
title: "Statistical Inference Course Project Part 1"
author: "Joe Schaal"
date: "Thursday, November 06, 2014"
output: html_document
---

This report discuss the results of a simulation of the **mean** value of **40** I.I.D. exponential random variables with lambda  $\lambda = 0.2$.  The simulation compares the distribution of **1000** trials with that of a normal distribution.

Mathematically, the **mean** $u$ and **standard deviation** $s$ of an exponential distribution with rate $\lambda$ is 
$1/\lambda$.   For this simulation,

```{r}
lambda <- 0.2
expMean <- 1/lambda
print(expMean)
```

and an individual trail is generated by 

```{r}
set.seed(5555)
simTrial <- mean(rexp(40,lambda))
print(simTrial)
```

Generating $n = 1000$ trials results in the histogram:
```{r}
trials <- 1000
sim <- replicate(n=trials,expr = mean(rexp(40,lambda)))
simMean <- mean(sim)
simHist <- hist(sim,breaks="Scott",main = "Histogram of Simulation")
yMax <- max(simHist$counts)
meanText <- paste("mean = ",as.character(format(mean(sim),digits=3)))
points(simMean,yMax,col="red",cex=2,pch=21,bg="red")
text(x = simMean+1 ,y = yMax, meanText)
```

The **mean** $u$ of this distribtuion is 
```{r}
simMean <- mean(sim)
print(simMean)
```

and the **standard deviation** $s$ is
```{r}
simSD <- sd(sim)
print(simSD)
```

This appears on the surface to be similiar to a normal distribution, centered near the theoretical mean of $\lambda = `r lambda`$.  If we overlay a normal distribution, with mean = `r simMean` and standard deviation `r simSD` we see that simulation distribution is centered very near the normal distribution. 
```{r}
simHist <- hist(sim,breaks="Scott",main = "Histogram of Simulation")
xfit<-seq(min(sim),max(sim),length=1000) 
yfit<-dnorm(xfit,mean=mean(sim),sd=sd(sim)) 
yfit <- yfit*diff(simHist$mids[1:2])*length(sim) 
lines(xfit, yfit, col="blue", lwd=2)
```

The pattern repeats itself in an additional trial of 1000:
```{r}
trials <- 1000
sim <- replicate(n=trials,expr = mean(rexp(40,lambda)))
simHist <- hist(sim,breaks="Scott",main = "Histogram of Simulation")
xfit<-seq(min(sim),max(sim),length=1000) 
yfit<-dnorm(xfit,mean=mean(sim),sd=sd(sim)) 
yfit <- yfit*diff(simHist$mids[1:2])*length(sim) 
lines(xfit, yfit, col="blue", lwd=2)
```

As well as in a larger trial with $n = 10000$:
```{r}
trials <- 10000
sim <- replicate(n=trials,expr = mean(rexp(40,lambda)))
simHist <- hist(sim,breaks="Scott",main = "Histogram of Simulation")
xfit<-seq(min(sim),max(sim),length=1000) 
yfit<-dnorm(xfit,mean=mean(sim),sd=sd(sim)) 
yfit <- yfit*diff(simHist$mids[1:2])*length(sim) 
lines(xfit, yfit, col="blue", lwd=2)
```