---
title: "Statistical Inference Course Project Part 1"
author: "Joe Schaal"
date: "Thursday, November 06, 2014"
output: pdf_document
---

This report discuss the results of a simulation of the **mean** value of **40** I.I.D. exponential random variables with lambda  $\lambda = 0.2$.  The simulation compares the distribution of **1000** trials with that of a normal distribution.

Mathematically, the **mean** $u$ and **standard deviation** $s$ of an exponential distribution with rate $\lambda$ is 
$\frac{1}{\lambda}$.   For this simulation,

```{r}
lambda <- 0.2
expMean <- 1/lambda
print(expMean)
```

and an individual trail is generated by 

```{r}
set.seed(5555)
n <- 40
simTrial <- mean(rexp(n,lambda))
print(simTrial)
```

Generating $trials = 1000$ trials results in the histogram:
```{r fig.width=4, fig.height=4}
trials <- 1000
sim <- apply(matrix(rexp(n*trials,lambda),trials),1,mean)
simMean <- mean(sim)
simSD <- sd(sim)
simHist <- hist(sim,breaks="Scott",main = "Histogram of Simulation")
yMax <- max(simHist$counts)
meanText <- paste("mean = ",as.character(format(mean(sim),digits=4)))
points(simMean,yMax,col="red",cex=2,pch=21,bg="red")
text(x = simMean+1.5 ,y = yMax, meanText)
expMean <- 1/lambda
expSD <- (1/lambda)/sqrt(n)
```

The **theoretical mean** of of the exponential distribution is $\frac{1}{\lambda} = `r expMean`$.   This compares favorably to the **actual mean** of $`r format(simMean,digits=4)`$ and the distribution is centered near its theoretical mean.   The theoretical variance of a sample mean for a sample of size `r n` (represented by its **standard deviaition $s$**) is $\frac{\sigma}{\sqrt{n}} = \frac{(1/\lambda)}{\sqrt{40}} = \frac{`r 1/lambda`}{`r format(sqrt(n),digits=4)`} = `r format((expSD),digits=4)`$.  This also compares favorably to the actual standard deviation of $`r format(simSD,digits=4)`$.

If we overlay a normal distribution, with the theoretical **mean** `r format(expMean,digits=4)` and **standard deviation** `r format(expSD,digits=4)` we see that normal distribution is very similar to the simulation distribution.
```{r fig.width=4, fig.height=4}
simHist <- hist(sim,breaks="Scott",main = "Histogram of Simulation")
xfit<-seq(min(sim),max(sim),length=1000) 
yfit<-dnorm(xfit,mean=expMean,sd=expSD) 
yfit <- yfit*diff(simHist$mids[1:2])*length(sim) 
lines(xfit, yfit, col="blue", lwd=2)
points(simMean,yMax,col="red",cex=2,pch=21,bg="red")
text(x = simMean+1.5 ,y = yMax, meanText)
```


The pattern repeats itself in an additional trial of 1000:
```{r}
sim <- apply(matrix(rexp(n*trials,lambda),trials),1,mean)
```
```{r echo=FALSE, fig.width=4, fig.height=4}
simHist <- hist(sim,breaks="Scott",main = "Histogram of Simulation")
xfit<-seq(min(sim),max(sim),length=1000) 
yfit<-dnorm(xfit,mean=expMean,sd=expSD) 
yfit <- yfit*diff(simHist$mids[1:2])*length(sim) 
lines(xfit, yfit, col="blue", lwd=2)
yMax <- max(simHist$counts)
meanText <- paste("mean = ",as.character(format(mean(sim),digits=4)))
points(simMean,yMax,col="red",cex=2,pch=21,bg="red")
text(x = simMean+1.5 ,y = yMax, meanText)
```

Here the **mean** of the sample is $`r format(mean(sim),digits=4)`$ and the **standard deviation** is $`r format(sd(sim),digits=4)`$.

As well as in a larger trial with $trials = 10000$:
```{r}
trials <- 10000
sim <- apply(matrix(rexp(n*trials,lambda),trials),1,mean)
```
```{r echo=FALSE, fig.width=4, fig.height=4}
simHist <- hist(sim,breaks="Scott",main = "Histogram of Simulation")
xfit<-seq(min(sim),max(sim),length=1000) 
yfit<-dnorm(xfit,mean=expMean,sd=expSD) 
yfit <- yfit*diff(simHist$mids[1:2])*length(sim) 
lines(xfit, yfit, col="blue", lwd=2)
yMax <- max(simHist$counts)
meanText <- paste("mean = ",as.character(format(mean(sim),digits=4)))
points(simMean,yMax,col="red",cex=2,pch=21,bg="red")
text(x = simMean+1.5 ,y = yMax, meanText)
```

Here the **mean** of the sample is $`r format(mean(sim),digits=4)`$ and the **standard deviation** is $`r format(var(sim),digits=4)`$.

This exercise shows that in general, when estimating the averages of a series of **I.I.D.** random variables from a common distribution, for sufficiently large $n$, the distribution is centered around the **expected value** of the population average,  has a **variance** that is proportional to the variance of the population:  $\frac{\sigma^2}{n}$, and approaches the distribution of a normal distribution with the equivalent mean and standard deviation.

This differs from a comparison against a sample of individual random variables from a population, which accordingly, will follow its inherent distribution.  For example for a standard uniform variable with range (0,1), the mean $u$ is $\frac{1}{2}(a+b) = 0.5$ and standard deviation $s = \sqrt{\sigma^2} =\sqrt{\frac{1}{12}(b-a)^2} = \sqrt{\frac{1}{12}} = `r format(sqrt(1/12),digits=4)`$.

```{r fig.width=4, fig.height=4}
trials <- 1000
sim <- runif(trials)
simHist <- hist(sim)
simMean <- mean(sim)
simSD <- sd(sim)
xfit<-seq(min(sim),max(sim),length=1000) 
yfit<-dnorm(xfit,mean=0.5,sd=sqrt(1/12))
yfit <- yfit*diff(simHist$mids[1:2])*length(sim) 
lines(xfit, yfit*.87, col="blue", lwd=2)
```

The **mean** $u = `r format(simMean,digits=4)`$ and **standard deviation**  $s = `r format(simSD,digits=4)`$ of the sample are very close to that of the underlying distribution but the distribution does not follow that of a normal distribution.   

However, if we instead sample the averages of a series of uniform random variables, such as sampling the average of $n = 40$ random variables, we see a very different pattern.  For the a sample of the mean of the population of uniform variables, we expect the mean to approach $\frac{1}{2}(a+b) = 0.5$ and the standard deviation to approach $s = \sqrt{\frac{\sigma^2}{n}} =\sqrt{\frac{1/12}{`r n`}} =  `r format(sqrt((1/12)/n),digits=4)`$
```{r fig.width=4, fig.height=4}
trials <- 1000
n <- 40
sim <- apply(matrix(runif(n*trials),trials),1,mean)
simHist <- hist(sim)
simMean <- mean(sim)
simSD <- sd(sim)
uniMean <- 1/2
uniSD <- sqrt((1/12)/n)
xfit<-seq(min(sim),max(sim),length=1000) 
yfit<-dnorm(xfit,mean=uniMean,sd=uniSD)
yfit <- yfit*diff(simHist$mids[1:2])*length(sim) 
lines(xfit, yfit*.87, col="blue", lwd=2)

```

In the simulation example, the **mean** of the simulation is $`r format(simMean,digits=4)`$ and the **standard deviation** is  $`r format(simSD,digits=4)`$.  This compares favorably with that of the theoretical mean and that of a similar normal distribution.