---
title: "Statistical Inference Course Project Part 1"
author: "Joe Schaal"
date: "Thursday, November 06, 2014"
output: html_document
---

This report discuss the results of a simulation of the **mean** value of **40** I.I.D. exponential random variables with lambda  $\lambda = 0.2$.  The simulation compares the distribution of **1000** trials with that of a normal distribution.

Mathematically, the **mean** $u$ and **standard deviation** $s$ of an exponential distribution with rate $\lambda$ is 
$1/\lambda$.   For this simulation,

```{r}
lambda <- 0.2
simMean <- 1/lambda
print(simMean)
```

and an individual trail is generated by 

```{r}
set.seed(5555)
simTrial <- mean(rexp(40,lambda))
simTrial
```

Generating $n = 1000$ trials results in the histogram:
```{r}
trials <- 1000
sim <- replicate(n=trials,expr = mean(rexp(n,lambda)))
simHist <- hist(sim,breaks="Scott")
```

The **mean** $u$ of this distribtuion is 
```{r}
simMean <- mean(sim)
print(simMean)
```

and the **standard deviation** %s% is
```{r}
simSD <- sd(sim)
print(simdSd)
```

This appears on the surface to be similiar to a normal distribution, with a **mean** 
