---
title: "Statistical Inference Course Project Part 1"
author: "Joe Schaal"
date: "Thursday, November 06, 2014"
output: pdf_document
---

``` {r initialization, echo=FALSE}
set.seed(5555)
n <- 40
trials <-1000
lambda <- 0.2
expMean <- 1/lambda
expSD <- (1/lambda)/sqrt(n)
figureCount <- 1

figureNumber <- function(name,figureCount) {
    figureCount <<- figureCount+1
    figureText <- paste("Fig",as.character(figureCount),":",name)
}
```
This report discuss the results of a simulation of the **mean** value of **`r n`** I.I.D. exponential random variables with lambda  $\lambda = $ **`r lambda`**.  The simulation compares the distribution of **`r trials`** trials with that of a normal distribution.  Mathematically, the **mean** $u$ and **standard deviation** $s$ of an exponential distribution with rate $\lambda$ is $1/\lambda$.  For this simulation, $u$ and $s$ are each equal to **`r expMean`**.  

Generating `r trials` trials results in the following histogram:
```{r figure_1}
sim1 <- apply(matrix(rexp(n*trials,lambda),trials),1,mean)
```
```{r figure_1_plot, echo=FALSE, fig.width=4, fig.height=4}
simMean <- mean(sim1)
simSD <- sd(sim1)
simHist <- hist(sim1,breaks="Scott",main = figureNumber("Histogram of Simulation 1",figureCount),xlab="")
yMax <- max(simHist$counts)
meanText <- paste("mean = ",as.character(round(simMean,4)))
sdText <- paste("sd = ",as.character(round(simSD,4)))
points(simMean,yMax,col="red",cex=2,pch=21,bg="red")
text(x = simMean+1.5 ,y = yMax, meanText)
text(x = simMean+1.5 ,y = yMax-8, sdText)
```

The **theorectical mean** of the distribution is **`r round(expMean,4)`** which compares favorably to the **simulation mean** of **`r round(simMean,4)`** and the distribution is centered near its theoretical mean.  The theoretical variance of a sample mean for a sample of size `r n` (represented by its **standard deviaition $s$**) is $\frac{\sigma}{\sqrt{n}} = \frac{(1/\lambda)}{\sqrt{40}} = \frac{`r 1/lambda`}{`r round(sqrt(n),4)`} = $**`r round(expSD,4)`**.  This also compares favorably to the **actual standard deviation** of **`r round(simSD,4)`**.

If we overlay a normal distribution, with the theoretical **mean** `r round(expMean,4)` and **standard deviation** `r round(expSD,4)` we see that normal distribution is very similar to the simulation distribution.

```{r figure_2, echo=FALSE, fig.width=4, fig.height=4}
simHist <- hist(sim1,breaks="Scott",main = figureNumber("Normal Distribution Overlay",figureCount),xlab = "")
xfit<-seq(min(sim1),max(sim1),length=1000) 
yfit<-dnorm(xfit,mean=expMean,sd=expSD) 
yfit <- yfit*diff(simHist$mids[1:2])*length(sim1) 
lines(xfit, yfit, col="blue", lwd=2)
```

This exercise shows that in general, when estimating the averages of a series of **I.I.D.** random variables from a common distribution, for sufficiently large $n$, the distribution is centered around the **expected value** of the population average,  has a **variance** that is proportional to the variance of the population:  $\frac{\sigma^2}{n}$, and approaches the distribution of a normal distribution with the equivalent mean and standard deviation.
```{r uniform, echo=FALSE}
a <- 0
b <- 1
unfMean <- 1/2*(a+b)
unfSD <- sqrt((1/12)*(b-a)^2)
unfSampleSD <- sqrt((unfSD^2)/n)
```
This differs from a comparison against a sample of individual random variables from a population, which accordingly, will follow its inherent distribution.  For example, for a standard uniform variable with range (`r a`,`r b`), the mean $u$ is $\frac{1}{2}(`r a`+`r b`) =$ **`r unfMean`** and standard deviation $s = \sqrt{\sigma^2} =\sqrt{\frac{1}{12}(`r b`-`r a`)^2} = $**`r round(unfSD,4)`**.

Simulating **`r trials`** of individual uniform distributions results in the following distribution:
```{r figure_3}
sim2 <- runif(trials)
```
```{r echo=FALSE,figure_3_plot, fig.width=4, fig.height=4}
simHist <- hist(sim2,breaks="Scott",main = figureNumber("Histogram of Simulation 2",figureCount),xlab="")
simMean <- mean(sim2)
simSD <- sd(sim2)
xfit<-seq(min(sim2),max(sim2),length=1000) 
yfit<-dnorm(xfit,mean=unfMean,sd=unfSD)
yfit <- yfit*diff(simHist$mids[1:2])*length(sim2) 
lines(xfit, yfit*.85, col="blue", lwd=2)
yMax <- max(simHist$counts)
meanText <- paste("mean = ",as.character(round(simMean,4)))
sdText <- paste("sd = ",as.character(round(simSD,4)))
points(simMean,yMax,col="red",cex=2,pch=21,bg="red")
mtext(text = meanText,side = 4,line=0)
mtext(sdText,4,line=1)
```

The **mean** $u$ = `r round(simMean,4)` and **standard deviation** $s$ = `r round(simSD,4)` of the sample are very close to that of the underlying distribution but the distribution does not follow that of a normal distribution.   

However, if we instead sample the **averages** of a series of uniform random variables, such as sampling the average of $n = `r n`$ random variables, we see a very different pattern.  For the  sample of the mean of the population of uniform variables, we expect the mean to approach $\frac{1}{2}(a+b) = `r unfMean`$ and the standard deviation to approach $s = \sqrt{\frac{\sigma^2}{n}} =\sqrt{\frac{(1/12)(`r b`-`r a`)^2}{`r n`}} =  `r round(sqrt((unfSD^2)/n),4)`$
```{r figure_4}
sim3 <- apply(matrix(runif(n*trials),trials),1,mean)
```
```{r echo=FALSE,fig.width=4, fig.height=4}
simHist <- hist(sim3,breaks="Scott",main = figureNumber("Histogram of Simulation 3",figureCount),xlab="")
simMean <- mean(sim3)
simSD <- sd(sim3)
xfit<-seq(min(sim3),max(sim3),length=1000) 
yfit<-dnorm(xfit,mean=unfMean,sd=unfSampleSD)
yfit <- yfit*diff(simHist$mids[1:2])*length(sim3) 
lines(xfit, yfit, col="blue", lwd=2)
yMax <- max(simHist$counts)
meanText <- paste("mean = ",as.character(round(simMean,4)))
sdText <- paste("sd = ",as.character(round(simSD,4)))
points(simMean,yMax,col="red",cex=2,pch=21,bg="red")
text(x = simMean+.08 ,y = yMax, meanText)
text(x = simMean+.08 ,y = yMax-14, sdText)

```

In the simulation example, the **mean** of the simulation is **`r format(simMean,digits=4)`** and the **standard deviation** is  **`r round(simSD,digits=4)`**.  This compares favorably with that of the theoretical mean and standard distribution, as well as that of a normal distribution with mean **`r unfMean`** and standard deviation **`r round(unfSampleSD,4)`**.

